---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tidyverse, quietly = TRUE)
library(sf, quietly = TRUE)
```

# london_liverpool_data

This repository holds spatial dataframes related to a paper... Below is an explanation of their sources.

## London

This dataframe contains polygons for the wards and boroughs of London. It contains 625 unique rows.

The data is available from the [London Datastore](https://data.london.gov.uk/) which was "created by the Greater London Authority (GLA) as a first step towards freeing London’s data".

This particular file appears as a collection of shapefiles in zipped form [here](https://data.london.gov.uk/dataset/statistical-gis-boundary-files-london%20as%20a%20zipped%20set%20of%20shapefiles%20called:%20statistical-gis-boundaries-london.zip).

They have been uploaded here as one `.rds` file.

```{r}

london <- readRDS("data/london.rds") |> 
  st_as_sf()

head(london)

```

```{r}
ggplot(london) + geom_sf() + theme_bw()
```

## River Thames

The dataframe for the river Thames in the Greater London area was downloaded from OpenStreetMap via a query in the [overpass-turbo](https://overpass-turbo.eu) wizard.

```{r}

thames <- readRDS("data/thames.rds") |> 
  st_as_sf() |> 
  select(-(17:74))

head(thames)

```

```{r}
ggplot(thames) + 
  geom_sf(colour="darkblue") + 
  theme_bw()
```

## Liverpool

The Liverpool dataset comes from the [Office of National Statistics](https://www.nomisweb.co.uk/home/census2001.asp) as presented by Francisco Rowe and Dani Arribas-Bel in their demonstration of techniques for hierarchical spatial regression. It is available in shapefile format at their [course repository](https://gdsl-ul.github.io/san/) for Spatial Modeling for Data Scientists, ENVS453, University of Liverpool.

The authors describe its contents as follows:

| row name | description                                                                       |
|------------|------------------------------------------------------------|
| oa_cd    | Output Area Code                                                                  |
| lsoa_cd  | Lower Super Output Area Code                                                      |
| msoa_cd  | Middle Super Output Area Code                                                     |
| lad_cd   | Local Authority District Code                                                     |
| ward_nm  | Ward Name                                                                         |
| dstrt_nm | District Name                                                                     |
| cnty_nm  | County Name                                                                       |
| cntry_nm | Country Name                                                                      |
| pop      | Population                                                                        |
| age_60   | Proportion of resident population aged 60+                                        |
| unemp    | Unemployment rate                                                                 |
| lat      | Latitude                                                                          |
| long     | Longitude                                                                         |
| males    | Percentage of male resident population                                            |
| lt_ill   | Percentage of population suffering from a long-term illness condition             |
| Bhealth  | Percentage of resident population reporting bad health                            |
| VBhealth | Percentage of resident population reporting very bad health                       |
| no_qual  | Percentage of resident population with no educational qualification               |
| manprof  | Percentage of resident population in high managerial and professional occupations |
| geometry | Geometry                                                                          |

: Source: 2011 UK Census, ONS: http://infuse.mimas.ac.uk/

```{r}

liverpool <- readRDS("data/liverpool.rds") |> 
  st_as_sf()

head(liverpool)

```

```{r}
ggplot(liverpool) + 
  geom_sf() + 
  theme_bw()
```

Some slight modifications were made to the original file such as making the administrative codes into factors, scaling the limiting long-term illness (LLTI) variable, and giving unemployment an unabbreviated name.

```{r, eval=FALSE}
 liverpool <- downloaded_file |>
  st_make_valid() |>
  st_transform(4326) |>
  mutate(
    oa_cd = factor(oa_cd),
    lsoa_cd = factor(lsoa_cd),
    msoa_cd = factor(msoa_cd),
    lt_illness = as.vector(scale(lt_ill))) |>
  rename(unemployment = unemp)
```

## River Mersey

The river Mersey dataset shows the mouth of the river in the Liverpool region and comes from the UK Government Environment Agency's [data services platform](https://environment.data.gov.uk/catchment-planning/WaterBody/GB531206908100)

```{r}

mersey <- readRDS("data/mersey.rds") |> 
  st_as_sf()

head(mersey)

```

```{r}
ggplot(mersey) + 
  geom_sf(fill="lightblue") + 
  theme_bw()
```

## Thames River Crossings

Information about the various crossing of the river Thames were scraped from [Wikipedia](https://en.wikipedia.org/wiki/List_of_crossings_of_the_River_Thames) using the following code. The dataset `crossing_roadped` contains crossings by road or which are available to pedestrians. This excludes things such as railway crossings, service or utility tunnels, gondola lifts, and flood barriers.

```{r}
### Scrape river-crossings data
library(rvest, quietly = TRUE)
library(httr, quietly = TRUE)
library(polite, quietly = TRUE)
library(janitor, quietly = TRUE)

url <- "https://en.wikipedia.org/wiki/List_of_crossings_of_the_River_Thames"

url_bow <- polite::bow(url)

ind_html <-
  polite::scrape(url_bow) %>%  # scrape web page
  rvest::html_nodes("table.wikitable") %>% # pull out specific table
  rvest::html_table(fill = TRUE) 

# having scraped data, a bit of a messy procedure below to tidy it up into sf dataframe...

names(ind_html) <- c(
  "North Sea to London",
  "East London",
  "Central London",
  "South West London",
  "London to Windsor",
  "Windsor to Reading",
  "Reading to Oxford",
  "Oxford to Cricklade",
  "Cricklade to the source"
)

# only want the crossings within greater london (east, central and west)
east <- 
  ind_html[[2]] |>  
  clean_names() |> 
  select(1:3)
names(east) <- c("crossing","type","coords")

central <- 
  ind_html[[3]] |>  
  clean_names() |> 
  select(1:3)
names(central) <- c("crossing","type","coords")

west <- 
  ind_html[[4]] |>  
  clean_names() |> 
  select(1:3)
names(west) <- c("crossing","type","coords")

crossings <- rbind(east,central,west) |> 
  data.frame() 

# dont want crossings such as "Utility tunnel[10]","Flood barrier","Gondola lift, transport route","Utility tunnel", "Ferry route"
# filter by road or pedestrian crossings
roadped <- which(str_detect(crossings$type,"Pedestrian | pedestrian | road | Road | Bridge | footbridge"))

crossings_roadped <- crossings[roadped,]

# more messing around to change from awkward wikipedia degree/minute/second structure...

# Extract content up to and including the letter "W" or "E"
crossings_roadped$coords <- sub("(.*W|E.*)\\b.*", "\\1", crossings_roadped$coords)

# Function to extract degrees, minutes, and seconds
extract_coords <- function(coord) {
  # Split the coordinates by space
  parts <- unlist(strsplit(coord, " "))
  
  # Extract degrees, minutes, and seconds for latitude
  lat_degrees <- as.numeric(gsub("[^0-9.-]", "", substr(parts[1], 1, regexpr("°", parts[1])-1)))
  lat_minutes <- as.numeric(gsub("[^0-9.-]", "", substr(parts[1], regexpr("°", parts[1])+1, regexpr("′", parts[1])-1)))
  lat_seconds <- as.numeric(gsub("[^0-9.-]", "", substr(parts[1], regexpr("′", parts[1])+1, regexpr("″", parts[1])-1)))
  
  # Extract degrees, minutes, and seconds for longitude
  lon_degrees <- as.numeric(gsub("[^0-9.-]", "", substr(parts[2], 1, regexpr("°", parts[2])-1)))
  lon_minutes <- as.numeric(gsub("[^0-9.-]", "", substr(parts[2], regexpr("°", parts[2])+1, regexpr("′", parts[2])-1)))
  lon_seconds <- as.numeric(gsub("[^0-9.-]", "", substr(parts[2], regexpr("′", parts[2])+1, regexpr("″", parts[2])-1)))
  
  return(c(lat_degrees, lat_minutes, lat_seconds, lon_degrees, lon_minutes, lon_seconds))
}

# Apply function to coords column
coords_split <- t(sapply(crossings_roadped$coords, extract_coords))

# Convert to data frame
coords_df <- as.data.frame(coords_split)

# Rename columns
names(coords_df) <- c("lat_degrees", "lat_minutes", "lat_seconds", "lon_degrees", "lon_minutes", "lon_seconds")

coords_df$latitude <- coords_df$lat_degrees + (coords_df$lat_minutes/60) + (coords_df$lat_seconds/3600)
coords_df$longitude <- coords_df$lon_degrees + (coords_df$lon_minutes/60) + (coords_df$lon_seconds/3600)

crossings_roadped$latitude <- coords_df$latitude
crossings_roadped$longitude <- coords_df$longitude

# multiply those west of Greenwich meridian by -1
crossings_roadped$longitude[which(!str_detect(crossings_roadped$coords, "E$"))] <- crossings_roadped$longitude[which(!str_detect(crossings_roadped$coords, "E$"))] * -1

# nudge Greenwich tunnel location to put it exactly on river
crossings_roadped$latitude[crossings_roadped$crossing == "Greenwich foot tunnel"] <- 51.485
crossings_roadped$longitude[crossings_roadped$crossing == "Greenwich foot tunnel"] <- -0.01

# nudge Woolwich tunnel location to put it exactly on river
crossings_roadped$latitude[crossings_roadped$crossing == "Woolwich foot tunnel"] <- 51.497

# finally, sf dataframe with same crs as thames and london...
crossings_roadped <- crossings_roadped |> st_as_sf(coords = c("longitude","latitude"), crs=4326)

```

```{r}
ggplot() + 
  geom_sf(data=thames, colour="darkblue") + 
  geom_sf(data=crossings_roadped, shape=7, size=5) + 
  theme_bw()
```
